---
title: "Ambient removal report"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    theme: "flatly"
    highlight: "haddock"
    toc: true
    toc_depth: 2
    toc_float: true
    df_print: "paged"
    self_contained: true
params:
  rmd: "ambient_removal.Rmd"
  sample_id: null
  droplet_qc: null
  libraries_qc: null
  outdir: null
  gex_matrix: null
  atac_matrix: null
  adt_matrix: null
  adt_filters: !r list("isotype_control" = NULL)
  control: null
  gex_ambient_removal: !r list("integrate_samples" = FALSE, "cluster_ndims" = 10, "cluster_k" = 20, "cluster_res" = 0.05, "cluster_markers" = NULL)
  adt_ambient_removal: !r list("integrate_samples" = FALSE, "cluster_ndims" = 10, "cluster_k" = 20, "cluster_res" = 0.2, "cluster_markers" = NULL)
---

```{r setup, include=FALSE}
# Set chunk options
knitr::opts_chunk$set(
  echo = FALSE,
  error = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "",
  results = "asis"
)

# Load packages
library(dplyr)
library(ggplot2)
library(Seurat)
library(decontX)

# Check if running within Snakemake pipeline
if (exists("snakemake")) {
  # if running within Snakemake pipeline
  log <- file(snakemake@log[[1]], open = "w")
  sink(log, type = "message") # write stderr to Snakemake log file

  snakemake@source("functions.R")

  unlockBinding("params", .GlobalEnv)
  params$sample_id <- snakemake@wildcards$sample
  params <- update.params(snakemake@params)
  lockBinding("params", .GlobalEnv)

  if (!dir.exists(dirname(snakemake@output[[1]]))) dir.create(dirname(snakemake@output[[1]]), recursive = TRUE)
} else {
  # if running in isolation
  source("functions.R")
}

# Replace path placeholders
unlockBinding("params", .GlobalEnv)
for (x in c("droplet_qc", "libraries_qc", "outdir", "gex_matrix", "atac_matrix", "adt_matrix")) {
  params[[x]] <- glue::glue(params[[x]], sample = params$sample_id)
  if (length(params[[x]]) == 0) params[[x]] <- NULL # override default value of glue::glue() for null input (character(0)) and return null output
}
lockBinding("params", .GlobalEnv)

# Check for missing required params
required <- c(
  "sample_id",
  "droplet_qc",
  "libraries_qc",
  "outdir"
)
is.missing <- sapply(params[required], is.null)
if (any(is.missing)) {
  names <- paste(names(params[required])[is.missing], collapse = ", ")
  stop(sprintf("The following required parameters are missing: %s", names))
}

# Initialise results list
counts <- list("filtered" = list(), "decontaminated" = list())
```

### Sample: `r params$sample_id`

# Overview

1. Load count matrices and subset to barcodes that passed libraries processing QC
2. Perform coarse clustering and run decontamination algorithm to remove ambient counts (GEX and ADT libraries only)
3. Output filtered and decontaminated count matrices in 10x HDF5 format

# Data loading

```{r load-libraries-qc-results}
cat("* **Libraries QC results:**", params$libraries_qc, "\n")
cells <- vroom::vroom(
  file.path(params$libraries_qc, "cell_barcodes.txt"),
  delim = "\t",
  col_names = FALSE,
  show_col_types = FALSE
) %>%
  pull(1)
```

```{r load-droplet-qc-results}
cat("* **Droplet processing QC results:**", params$droplet_qc, "\n")
droplet.qc.res <- vroom::vroom(
  file.path(params$droplet_qc, "metadata.tsv"),
  delim = "\t",
  col_names = TRUE,
  show_col_types = FALSE
) %>%
  mutate(Sample = factor(Sample, levels = sort(unique(Sample)), ordered = TRUE)) %>%
  as.data.frame()
rownames(droplet.qc.res) <- droplet.qc.res$Barcode
droplet.qc.res <- droplet.qc.res[cells, ]

empty <- vroom::vroom(
  file.path(params$droplet_qc, "empty_droplet_barcodes.txt"),
  delim = "\t",
  col_names = FALSE,
  show_col_types = FALSE
) %>%
  pull(1)
```

```{r load-gex-matrix, eval=!is.null(params$gex_matrix)}
cat("* **GEX matrix:**", params$gex_matrix, "\n")
gex.mat <- lapply(
  list("cells" = cells, "background" = empty),
  get.10x.matrix,
  file = params$gex_matrix,
  type = "Gene Expression"
)
```

```{r load-atac-matrix, eval=!is.null(params$atac_matrix)}
cat("* **ATAC matrix:**", params$atac_matrix, "\n")
atac.mat <- get.10x.matrix(params$atac_matrix, cells = cells, type = "Peaks")
counts <- lapply(
  counts,
  function(list, mat) {
    list[["Peaks"]] <- mat
    return(list)
  },
  mat = atac.mat
)
```

```{r load-adt-matrix, eval=!is.null(params$adt_matrix)}
cat("* **ADT matrix:**", params$adt_matrix, "\n")
if (params$adt_matrix == params$gex_matrix) {
  adt.mat <- lapply(
    list("cells" = cells, "background" = empty),
    get.10x.matrix,
    file = params$adt_matrix,
    type = "Antibody Capture"
  ) %>%
    lapply(function(mat) mat[grep(pattern = "^ADT_", rownames(mat)), ])
} else {
  adt.mat <- lapply(
    list("cells" = cells, "background" = empty),
    get.barcounter.matrix,
    file = params$adt_matrix,
    features.pattern = "^ADT_"
  )
}

# Remove isotype control antibodies
if (!is.null(params$adt_filters$isotype_control)) {
  adt.mat <- lapply(
    adt.mat,
    function(mat, isotype.control) mat[grep(pattern = isotype.control, rownames(mat), invert = TRUE), ],
    isotype.control = params$adt_filters$isotype_control
  )
}
```

# Data processing

`r if(!is.null(params$gex_matrix)) "## GEX ambient removal {.tabset}"`

```{r gex-ambient-removal, eval=!is.null(params$gex_matrix), include=FALSE}
seu <- CreateSeuratObject(
  counts = gex.mat$cells,
  assay = "RNA",
  meta.data = droplet.qc.res
)

if (params$gex_ambient_removal$integrate_samples) {
  if (!is.null(params$control)) seu[["donor_type"]] <- ifelse(seu$Sample == params$control, "Control", "MS")
  seu.list <- SplitObject(seu, split.by = ifelse(is.null(params$control), "Sample", "donor_type"))
  seu.list <- lapply(
    seu.list,
    function(obj) {
      obj <- obj %>%
        NormalizeData() %>%
        FindVariableFeatures()
      return(obj)
    }
  )
  features <- SelectIntegrationFeatures(seu.list)
  seu.list <- lapply(
    seu.list,
    function(obj) {
      obj <- obj %>%
        ScaleData(features = features) %>%
        RunPCA(features = features)
    }
  )
  anchors <- FindIntegrationAnchors(
    seu.list,
    anchor.features = features,
    reduction = "rpca"
  )
  seu <- IntegrateData(anchors)
  seu[["RNA"]] <- JoinLayers(seu[["RNA"]])
  DefaultAssay(seu) <- "integrated"
} else {
  seu <- seu %>%
    NormalizeData() %>%
    FindVariableFeatures()
}

seu <- seu %>%
  ScaleData() %>%
  RunPCA() %>%
  RunUMAP(dims = seq.int(1, params$gex_ambient_removal$cluster_ndims)) %>%
  FindNeighbors(
    dims = seq.int(1, params$gex_ambient_removal$cluster_ndims),
    k.param = params$gex_ambient_removal$cluster_k
  ) %>%
  FindClusters(resolution = params$gex_ambient_removal$cluster_res)

decont.gex <- decontX(gex.mat$cells, background = gex.mat$background, z = Idents(seu))
seu[["decontRNA"]] <- CreateAssayObject(counts = round(decont.gex$decontXcounts))

counts <- Map(
  function(list, assay, obj = seu) {
    list[["Gene Expression"]] <- GetAssayData(obj, assay = assay, layer = "counts")
    return(list)
  },
  list = counts,
  assay = c("RNA", "decontRNA")
)
```

```{r gex-ambient-removal-samples, eval=!is.null(params$gex_matrix)}
cat("### Samples", "\n\n")
UMAPPlot(seu, group.by = "Sample") +
  labs(title = "Samples") +
  theme_minimal()
```

```{r gex-ambient-removal-clusters, eval=!is.null(params$gex_matrix)}
cat("### Clusters", "\n\n")
UMAPPlot(seu, group.by = "ident") +
  labs(title = "Clusters") +
  theme_minimal()
```

```{r gex-ambient-removal-cluster-markers, fig.width=7, fig.height=2.5*ceiling(length(params$gex_ambient_removal$cluster_markers)/2), eval=all(!is.null(params$gex_matrix), !is.null(params$gex_ambient_removal$cluster_markers))}
cat("### Cluster markers", "\n\n")
plots <- FeaturePlot(
  seu,
  features = ensembldb::mapIds(EnsDb.Hsapiens.v86::EnsDb.Hsapiens.v86, keys = params$gex_ambient_removal$cluster_markers, keytype = "SYMBOL", columns = c("SYMBOL", "GENEID")),
  combine = FALSE
) %>%
  Map(
    function(x, title) {
      x +
        labs(title = title) +
        theme_minimal() +
        theme(legend.position = "none")
    },
    .,
    params$gex_ambient_removal$cluster_markers
  )
cowplot::plot_grid(plotlist = plots, ncol = 2)
```

```{r gex-ambient-removal-diagnostics, eval=!is.null(params$gex_matrix)}
cat("### Diagnostics", "\n\n")
cat("* **Pre-clustering sample integration:**", ifelse(params$gex_ambient_removal$integrate_samples, "Yes", "No"), "\n")
cat("* **Graph principal components:**", params$gex_ambient_removal$cluster_ndims, "\n")
cat("* **Graph nearest neighbours (k):**", params$gex_ambient_removal$cluster_k, "\n")
cat("* **Cluster resolution:**", params$gex_ambient_removal$cluster_res, "\n")
```

`r if(!is.null(params$adt_matrix)) "## ADT ambient removal {.tabset}"`

```{r adt-ambient-removal, eval=!is.null(params$adt_matrix), include=FALSE}
seu <- CreateSeuratObject(
  counts = adt.mat$cells,
  assay = "ADT",
  meta.data = droplet.qc.res
)
VariableFeatures(seu) <- rownames(seu)

if (params$adt_ambient_removal$integrate_samples) {
  if (!is.null(params$control)) seu[["donor_type"]] <- ifelse(seu$Sample == params$control, "Control", "MS")
  seu.list <- SplitObject(seu, split.by = ifelse(is.null(params$control), "Sample", "donor_type"))
  seu.list <- lapply(
    seu.list,
    function(obj) {
      obj <- obj %>%
        NormalizeData(normalization.method = "CLR", margin = 2) %>%
        ScaleData() %>%
        RunPCA()
      return(obj)
    }
  )
  anchors <- FindIntegrationAnchors(
    seu.list,
    anchor.features = features,
    reduction = "rpca"
  )
  seu <- IntegrateData(anchors)
  seu[["ADT"]] <- JoinLayers(seu[["ADT"]])
  DefaultAssay(seu) <- "integrated"
} else {
  seu <- seu %>%
    NormalizeData(normalization.method = "CLR", margin = 2)
}

seu <- seu %>%
  ScaleData() %>%
  RunPCA() %>%
  RunUMAP(dims = seq.int(1, params$adt_ambient_removal$cluster_ndims)) %>%
  FindNeighbors(
    dims = seq.int(1, params$adt_ambient_removal$cluster_ndims),
    k.param = params$adt_ambient_removal$cluster_k
  ) %>%
  FindClusters(resolution = params$adt_ambient_removal$cluster_res)

decont.adt <- decontPro(adt.mat$cells, cell_type = as.integer(Idents(seu), delta_sd = 5e-5, background_sd = 5e-6))
seu[["decontADT"]] <- CreateAssayObject(counts = round(decont.adt$decontaminated_counts))

counts <- Map(
  function(list, assay, obj = seu) {
    list[["Antibody Capture"]] <- GetAssayData(obj, assay = assay, layer = "counts")
    return(list)
  },
  list = counts,
  assay = c("ADT", "decontADT")
)
```

```{r adt-ambient-removal-samples, eval=!is.null(params$adt_matrix)}
cat("### Samples", "\n\n")
UMAPPlot(seu, group.by = "Sample") +
  labs(title = "Samples") +
  theme_minimal()
```

```{r adt-ambient-removal-clusters, eval=!is.null(params$adt_matrix)}
cat("### Clusters", "\n\n")
UMAPPlot(seu, group.by = "ident") +
  labs(title = "Clusters") +
  theme_minimal()
```

```{r adt-ambient-removal-cluster-markers, fig.width=7, fig.height=2.5*ceiling(length(params$adt_ambient_removal$cluster_markers)/2), eval=all(!is.null(params$adt_matrix), !is.null(params$adt_ambient_removal$cluster_markers))}
cat("### Cluster markers", "\n\n")
plots <- FeaturePlot(
  seu,
  features = paste("ADT", params$adt_ambient_removal$cluster_markers, sep = "-"),
  combine = FALSE
) %>%
  lapply(
    function(x) {
      x +
        theme_minimal() +
        theme(legend.position = "none")
    }
  )
cowplot::plot_grid(plotlist = plots, ncol = 2)
```

```{r adt-ambient-removal-decontaminated-counts, fig.width=3.5*round(sqrt(length(params$adt_ambient_removal$cluster_markers))), fig.height=1.5*ceiling(length(params$adt_ambient_removal$cluster_markers)/round(sqrt(length(params$adt_ambient_removal$cluster_markers)))), eval=all(!is.null(params$adt_matrix), !is.null(params$adt_ambient_removal$cluster_markers))}
cat("### Decontaminated counts", "\n\n")
plotDensity(
  GetAssayData(seu, assay = "ADT", layer = "counts"),
  GetAssayData(seu, assay = "decontADT", layer = "counts"),
  paste("ADT", params$adt_ambient_removal$cluster_markers, sep = "-")
)
```

```{r adt-ambient-removal-diagnostics, eval=!is.null(params$adt_matrix)}
cat("### Diagnostics", "\n\n")
cat("* **Pre-clustering sample integration:**", ifelse(params$adt_ambient_removal$integrate_samples, "Yes", "No"), "\n")
cat("* **Graph principal components:**", params$adt_ambient_removal$cluster_ndims, "\n")
cat("* **Graph nearest neighbours (k):**", params$adt_ambient_removal$cluster_k, "\n")
cat("* **Cluster resolution:**", params$adt_ambient_removal$cluster_res, "\n")
```

# Output

```{r output}
if (!dir.exists(params$outdir)) dir.create(params$outdir, recursive = TRUE)
for (type in names(counts)) {
  feature.counts <- sapply(counts[[type]], nrow)
  DropletUtils::write10xCounts(
    path = file.path(params$outdir, sprintf("%s_matrix.h5", type)),
    x = Reduce(rbind, counts[[type]]),
    gene.type = rep(names(feature.counts), times = feature.counts),
    genome = "GRCh38",
    version = "3",
    overwrite = TRUE
  )
}

cat("* **Output path:**", params$outdir, "\n")
```

# Session Info

```{r session-info, results="markup"}
sessionInfo()
```

# Source
This report was automatically generated from an R Markdown script (<a download="ambient_removal.Rmd" href="`r base64enc::dataURI(file = params$rmd, mime = 'text/rmd', encoding = 'base64')`">download source</a>).
