---
title: "Droplet processing QC report"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    theme: "flatly"
    highlight: "haddock"
    toc: true
    toc_depth: 2
    toc_float: true
    df_print: "paged"
    self_contained: true
params:
  rmd: "droplet_qc.Rmd"
  sample_id: NULL
  metadata: NULL
  output: NULL
  gex_matrix: NULL
  hto_matrix: NULL
  emptydrops_lower: 100
  emptydrops_niters: 10000
  demuxmix_model: "auto"
  demuxmix_pAcpt: 0.9
  multiplet_projection: "TSNE"
---

```{r setup, include=FALSE}
# Set chunk options
knitr::opts_chunk$set(
  echo = FALSE,
  error = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "",
  results = "asis"
)

# Load packages
library(dplyr)
library(ggplot2)
library(plotly)
library(DropletUtils)
library(demuxmix)
library(scuttle)
library(scater)
library(scran)
library(scDblFinder)

# Check if running within Snakemake pipeline
if (exists("snakemake")) {
  # if running within Snakemake pipeline
  log <- file(snakemake@log[[1]], open = "w")
  sink(log, type = "message") # write stderr to Snakemake log file

  snakemake@source("functions.R")

  unlockBinding("params", .GlobalEnv)
  params$sample_id <- snakemake@wildcards$sample
  params <- update.params(snakemake@params)
  lockBinding("params", .GlobalEnv)

  if (!dir.exists(dirname(snakemake@output[[1]]))) dir.create(dirname(snakemake@output[[1]]), recursive = TRUE)
} else {
  # if running in isolation
  source("functions.R")
}

# Replace path placeholders
unlockBinding("params", .GlobalEnv)
for (x in c("output", "gex_matrix", "hto_matrix")) {
  params[[x]] <- glue::glue(params[[x]], sample = params$sample_id)
  if (length(params[[x]]) == 0) params[[x]] <- NULL # override default value of glue::glue() for null input (character(0)) and return null output
}
lockBinding("params", .GlobalEnv)

# Check for missing required params
required <- c(
  "sample_id",
  "metadata",
  "output",
  "gex_matrix",
  "hto_matrix"
)
is.missing <- sapply(params[required], is.null)
if (any(is.missing)) {
  names <- paste(names(params[required])[is.missing], collapse = ", ")
  stop(sprintf("The following required parameters are missing: %s", names))
}
```

### Sample: `r params$sample_id`

# Overview

1. Load GEX and HTO count matrices
2. Run cell calling algorithm on GEX count matrix to filter empty droplet barcodes
3. Run HTO demultiplexing algorithm to classify barcodes as 'singlet' (with assigned HTO), 'multiplet', 'negative' or 'uncertain'
4. Run multiplet calling algorithm to predict intra-sample multiplets based on known inter-sample multiplets
5. Output CSV file containing cell metadata for barcodes passing QC filters

# Data loading

```{r load-metadata}
cat("* **Metadata:**", "\n")
cat("```", params$metadata, "```", sep = "\n")
md <- yaml::yaml.load(params$metadata)
```

```{r load-gex-matrix}
cat("* **GEX matrix:**", params$gex_matrix, "\n")
gex.mat <- get.10x.matrix(params$gex_matrix, type = "Gene Expression")
gex.mat <- gex.mat[, colSums(gex.mat) > 0] # remove cell barcodes with zero counts
```

```{r load-hto-matrix}
cat("* **HTO matrix:**", params$hto_matrix, "\n")
if (params$hto_matrix == params$gex_matrix) {
  hto.mat <- get.10x.matrix(params$hto_matrix, type = "Antibody Capture")
} else {
  hto.mat <- get.barcounter.matrix(params$hto_matrix, features.pattern = "^HTO_")
}
wl <- names(md)
hto.mat <- hto.mat[match(wl, rownames(hto.mat)), ]
```

# Data processing

```{r cell-calling}
cat("## Cell calling {.tabset}", "\n\n")
set.seed(100)
ed.res <- emptyDrops(gex.mat, lower = params$emptydrops_lower, niters = params$emptydrops_niters, test.ambient = TRUE)
ed.stats <- list(
  cells = sum(ed.res$FDR <= 0.001, na.rm = TRUE),
  empty = sum(ed.res$FDR >= 0.001, na.rm = TRUE),
  background = sum(is.na(ed.res$FDR <= 0.001)),
  limited = sum(ed.res$Limited[which(ed.res$FDR >= 0.001)]) > 0
)
bcrank.res <- barcodeRanks(gex.mat)
gex.mat <- gex.mat[, which(ed.res$FDR <= 0.001)]
bc.common <- intersect(colnames(gex.mat), colnames(hto.mat))
gex.mat <- gex.mat[, bc.common]
hto.mat <- hto.mat[, bc.common]
```

```{r cell-calling-classification}
cat("### Classification", "\n\n")
cat("* **Cell-containing droplets:**", ed.stats$cells %>% format(big.mark = ","), "\n")
cat("* **Empty droplets:**", ed.stats$empty %>% format(big.mark = ","), "\n")
cat("* **Background droplets:**", ed.stats$background %>% format(big.mark = ","), "\n")
cat("\n\n")
ggplotly(
  bcrank.plot(
    bcrank.res,
    ed.res,
    title = "Barcode rank plot",
    x.lab = "Rank",
    y.lab = "Total UMI count",
    colour.lab = "Fraction cells",
    log = TRUE
  ) +
    scale_colour_gradient2(low = "#d82526", mid = "#ffc156", high = "#69b764", midpoint = 0.5)
)
```

```{r cell-calling-diagnostics}
cat("### Diagnostics", "\n\n")
cat("* **Background UMI threshold:**", params$emptydrops_lower %>% format(big.mark = ","), "\n")
cat("* **Monte Carlo simulation iterations:**", params$emptydrops_niters %>% format(big.mark = ","), "\n")
cat("* **Limited:**", ifelse(ed.stats$limited, "Yes", "No"), "\n")
cat("\n\n")
ggplotly(
  pvalue.plot(
    ed.res,
    binwidth = 0.01,
    title = "Monte Carlo p-value distribution for background barcodes",
    x.lab = "p-value"
  )
)
```

```{r hto-demultiplexing}
cat("## HTO demultiplexing {.tabset}", "\n\n")
set.seed(100)
dmm <- demuxmix(
  hto = as.matrix(hto.mat),
  rna = colSums(gex.mat),
  model = params$demuxmix_model,
  pAcpt = params$demuxmix_pAcpt^length(wl)
)
models <- lapply(dmm@models, function(x) class(x))
demux.res <- dmm %>%
  dmmClassify() %>%
  mutate(Sample = if_else(grepl(",", HTO), "multiplet", HTO))
demux.stats <- demux.res %>%
  as.data.frame() %>%
  group_by(Type) %>%
  summarise(count = n())
demux.stats <- demux.stats$count %>%
  as.list() %>%
  setNames(demux.stats$Type)
```

```{r hto-demultiplexing-classification}
cat("### Classification", "\n\n")
ggplotly(
  bar.plot(
    data = demux.res %>% mutate(Type = factor(Type, levels = c("singlet", "multiplet", "negative", "uncertain"))),
    pos = Type,
    fill = Sample,
    position = position_stack(reverse = TRUE),
    title = "Sample assignment",
    pos.lab = "Type",
    fill.lab = "Sample"
  ),
  tooltip = c("fill", "x")
)
```

```{r hto-demultiplexing-raw-counts, fig.width=7, fig.height=length(wl)}
cat("### Raw HTO counts", "\n\n")
logcounts.plot(
  hto.mat,
  scale = 0.95,
  colour = NA,
  fill = "#595959",
  title = "Raw HTO counts"
)
```

```{r hto-demultiplexing-demux-counts, fig.width=7, fig.height=3*ceiling(length(wl)/2)}
cat("### Demultiplexed HTO counts", "\n\n")
logcounts.plot(
  hto.mat,
  samples = demux.res$Sample,
  scale = 0.95,
  colour = NA,
  show.legend = FALSE,
  title = "Demultiplexed HTO counts"
)
```

```{r hto-demultiplexing-mixture-components, fig.width=7, fig.height=2.5}
cat("### Mixture components", "\n\n")
for (x in wl) {
  plot <- plotDmmHistogram(dmm, hto = x) +
    labs(title = x, x = "UMI count") +
    theme_minimal()
  print(plot)
}
```

```{r hto-demultiplexing-diagnostics}
cat("### Diagnostics", "\n\n")
cat("* **Mode:**", params$demuxmix_model, "\n")
cat("* **Chosen model:**", "\n")
cat(sprintf("  + **%s:** %s", wl, models[wl]) %>% paste(collapse = "\n"), "\n")
cat("* **Acceptance probability:**", params$demuxmix_pAcpt^length(wl) %>% round(2), paste0("(", params$demuxmix_pAcpt %>% round(2), " per HTO)"), "\n")
cat("* **Fraction overlap between mixture components:**", "\n")
cat(sprintf("  + **%s:** %s", wl, round(dmmOverlap(dmm)[wl], 3)) %>% paste(collapse = "\n"), "\n")
```

```{r multiplet-calling}
cat("## Multiplet calling {.tabset}", "\n\n")
set.seed(100)
sce <- SingleCellExperiment(assays = list(counts = gex.mat[, demux.res$Type %in% c("singlet", "multiplet")])) %>%
  add.cell.metadata(demux.res[demux.res$Type %in% c("singlet", "multiplet"), "Sample", drop = FALSE]) %>%
  logNormCounts()
dec <- modelGeneVar(sce)
hvg <- getTopHVGs(dec, n = 1000)
sce <- sce %>%
  runPCA(subset_row = hvg, ncomponents = 20)
rd.res <- recoverDoublets(
  sce,
  use.dimred = "PCA",
  doublets = which(sce$Sample == "multiplet"),
  samples = table(sce$Sample) %>% head(-1) # remove count for multiplets
)
rownames(rd.res) <- colnames(sce)
rd.stats <- list(
  known = sum(rd.res$known),
  predicted = sum(rd.res$predicted),
  singlets = sum(!(rd.res$known | rd.res$predicted))
)
sce <- sce %>% add.cell.metadata(rd.res)
```

```{r multiplet-calling-classification}
cat("### Classification", "\n\n")
cat("* **Known (inter-sample) multiplets:**", rd.stats$known %>% format(big.mark = ","), "\n")
cat("* **Predicted (intra-sample) multiplets:**", rd.stats$predicted %>% format(big.mark = ","), "\n")
cat("\n\n")
ggplotly(
  multiplet.plot(
    sce,
    rd.res,
    size = 0.1,
    projection = "TSNE",
    use.dimred = "PCA",
    random.seed = 100
  ),
  tooltip = "colour"
)
```

# Output

```{r output}
out <- colData(sce)[!sce$known & !sce$predicted, "Sample", drop = FALSE] %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Barcode") %>%
  mutate(
    Donor = stringr::str_split_i(params$sample_id, pattern = "-", i = 1),
    Hash = stringr::str_split_i(params$sample_id, pattern = "-", i = 2),
    Sample = sapply(Sample, function(x) md[[x]]$sample)
  )
if (!dir.exists(dirname(params$output))) dir.create(dirname(params$output), recursive = TRUE)
vroom::vroom_write(
  out,
  file = params$output,
  delim = ","
)

cat("* **Total barcodes passed:**", rd.stats$singlets %>% format(big.mark = ","), "\n")
cat("* **Total barcodes excluded:**", sum(ed.stats$empty, ed.stats$background, demux.stats$negative, demux.stats$uncertain, rd.stats$known, rd.stats$predicted) %>% format(big.mark = ","), "\n")
cat("  + **Empty droplets:**", sum(ed.stats$empty, ed.stats$background) %>% format(big.mark = ","), "\n")
cat("  + **Negative/uncertain HTO:**", sum(demux.stats$negative, demux.stats$uncertain) %>% format(big.mark = ","), "\n")
cat("  + **Multiplets:**", sum(rd.stats$known, rd.stats$predicted) %>% format(big.mark = ","), "\n")
cat("\n\n")
out
```

# Session Info

```{r session-info, results="markup"}
sessionInfo()
```

# Source
This report was automatically generated from an R Markdown script (<a download="droplet_qc.Rmd" href="`r base64enc::dataURI(file = params$rmd, mime = 'text/rmd', encoding = 'base64')`">download source</a>).
