---
title: Droplet processing QC report
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    theme: flatly
    highlight: haddock
    toc: true
    toc_depth: 2
    toc_float: true
    df_print: paged
    self_contained: true
params:
  rmd: droplet_qc.Rmd
  sample_id: null
  metadata: null
  outdir: null
  gex_matrix: null
  atac_matrix: null
  adt_matrix: null
  hto_matrix: null
  pre_filtered: false
  cell_calling_algorithm: emptydrops
  emptydrops_umi_min: 500
  emptydrops_umi_min_frac_median: 0.01
  emptydrops_cand_max_n: 20000
  emptydrops_ind_min: 45000
  emptydrops_ind_max: 90000
  emptydrops_niters: 10000
  joint_modalities:
    - gex
    - atac
    - adt
  joint_ordmag_quantile: 0.99
  joint_ordmag_ratio: 10
  demuxmix_model: auto
  demuxmix_pAcpt: 0.9
  scdblfinder_clusters: true
  scdblfinder_dbr_sd: 1
  multiplet_projection: TSNE
---

```{r setup, include=FALSE}
# Set chunk options
knitr::opts_chunk$set(
  echo = FALSE,
  error = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "",
  results = "asis"
)

# Load packages
library(dplyr)
library(ggplot2)
library(plotly)
library(DropletUtils)
library(demuxmix)
library(scuttle)
library(scater)
library(scran)
library(scDblFinder)

# Check if running within Snakemake pipeline
if (exists("snakemake")) {
  # if running within Snakemake pipeline
  log <- file(snakemake@log[[1]], open = "w")
  sink(log, type = "message") # write stderr to Snakemake log file

  snakemake@source("functions.R")

  unlockBinding("params", .GlobalEnv)
  params$sample_id <- snakemake@wildcards$sample
  params <- update.params(snakemake@params)
  lockBinding("params", .GlobalEnv)

  if (!dir.exists(dirname(snakemake@output[[1]]))) dir.create(dirname(snakemake@output[[1]]), recursive = TRUE)
} else {
  # if running in isolation
  source("functions.R")
}

# Replace path placeholders
unlockBinding("params", .GlobalEnv)
for (x in c("outdir", "gex_matrix", "atac_matrix", "adt_matrix", "hto_matrix")) {
  params[[x]] <- glue::glue(params[[x]], sample = params$sample_id)
  if (length(params[[x]]) == 0) params[[x]] <- NULL # override default value of glue::glue() for null input (character(0)) and return null output
}
lockBinding("params", .GlobalEnv)

# Check for missing required params
required <- c(
  "sample_id",
  "metadata",
  "outdir",
  "gex_matrix"
)
is.missing <- sapply(params[required], is.null)
if (any(is.missing)) {
  names <- paste(names(params[required])[is.missing], collapse = ", ")
  stop(sprintf("The following required parameters are missing: %s", names))
}
params$cell_calling_algorithm <- match.arg(tolower(params$cell_calling_algorithm), choices = c("emptydrops", "joint"))
if (!params$pre_filtered && params$cell_calling_algorithm == "joint" && is.null(params$atac_matrix)) {
  stop("Missing parameter 'atac_matrix' is required for 'joint' cell calling algorithm.")
}
params$joint_modalities <- match.arg(tolower(params$joint_modalities), choices = c("gex", "atac", "adt"), several.ok = TRUE)

bc <- list()
```

### Sample: `r params$sample_id`

# Overview

1. Load GEX and HTO count matrices
2. Run cell calling algorithm to filter empty droplet barcodes (if matrices are not pre-filtered to only cell-containing barcodes)
3. Run HTO demultiplexing algorithm to classify barcodes as 'singlet' (with assigned HTO), 'multiplet', 'negative' or 'uncertain' (if HTO matrix provided)
4. Run multiplet calling algorithm
  * Predict intra-sample multiplets based on known inter-sample multiplets (if HTO matrix provided)
  * Predict multiplets based on simulated doublets (if HTO matrix not provided)
5. Output TSV file containing cell metadata for barcodes passing QC filters and text file containing list of empty droplet barcodes (if matrices are not pre-filtered to only cell-containing barcodes)

# Data loading

```{r load-metadata}
cat("* **Metadata:**", "\n")
cat("```", params$metadata, "```", sep = "\n")
md <- yaml::yaml.load(params$metadata)
```

```{r load-gex-matrix}
cat("* **GEX matrix:**", params$gex_matrix, "\n")
gex.mat <- get.10x.matrix(params$gex_matrix, type = "Gene Expression")
gex.mat <- gex.mat[, colSums(gex.mat) > 0] # remove cell barcodes with zero counts
bc$gex <- colnames(gex.mat)
```

```{r load-atac-matrix, eval=!is.null(params$atac_matrix)}
cat("* **ATAC matrix:**", params$atac_matrix, "\n")
atac.mat <- get.10x.matrix(params$atac_matrix, type = "Peaks")
atac.mat <- atac.mat[, colSums(atac.mat) > 0] # remove cell barcodes with zero counts
bc$atac <- colnames(atac.mat)
```

```{r load-adt-matrix, eval=!is.null(params$adt_matrix)}
cat("* **ADT matrix:**", params$adt_matrix, "\n")
if (params$adt_matrix == params$gex_matrix) {
  adt.mat <- get.10x.matrix(params$adt_matrix, type = "Antibody Capture")
} else {
  adt.mat <- get.barcounter.matrix(params$adt_matrix, features.pattern = "^ADT_")
}
adt.mat <- adt.mat[, colSums(adt.mat) > 0] # remove cell barcodes with zero counts
bc$adt <- colnames(adt.mat)
```

```{r load-hto-matrix, eval=!is.null(params$hto_matrix)}
cat("* **HTO matrix:**", params$hto_matrix, "\n")
if (params$hto_matrix == params$gex_matrix) {
  hto.mat <- get.10x.matrix(params$hto_matrix, type = "Antibody Capture")
} else {
  hto.mat <- get.barcounter.matrix(params$hto_matrix, features.pattern = "^HTO_")
}
wl <- names(md)
hto.mat <- hto.mat[match(wl, rownames(hto.mat)), ]
hto.mat <- hto.mat[, colSums(hto.mat) > 0] # remove cell barcodes with zero counts
bc$hto <- colnames(hto.mat)
```

# Data processing

```{r no-cell-calling, eval=params$pre_filtered}
bc$gex <- colnames(gex.mat)
bc.cells <- Reduce(intersect, bc)
gex.mat <- gex.mat[, bc.cells]
```

```{r cell-calling, eval=!params$pre_filtered}
cat("## Cell calling {.tabset}", "\n\n")
modalities <- names(bc)[names(bc) != "hto"]
total.counts <- lapply(
  modalities,
  function(x) {
    data.frame(colSums(get(paste(x, "mat", sep = ".")))) %>%
      setNames(x) %>%
      tibble::rownames_to_column("barcode")
  }
) %>%
  Reduce(
    x = .,
    f = function(x, y) full_join(x, y, by = "barcode")
  ) %>%
  filter(if_all(everything(), function(x) !is.na(x)))

if (params$cell_calling_algorithm == "joint") {
  cc.res <- joint.cell.caller(
    matrix.list = lapply(params$joint_modalities[match(modalities, params$joint_modalities)], function(x) get(paste(x, "mat", sep = "."))),
    ordmag.quantile = params$joint_ordmag_quantile,
    ordmag.ratio = params$joint_ordmag_ratio
  )
  bc.empty <- bc$gex[!(bc$gex %in% cc.res)]
  cc.stats <- list(
    cells = length(cc.res),
    empty = length(bc.empty)
  )
  bc$gex <- bc$gex[bc$gex %in% cc.res]
} else {
  set.seed(100)
  cc.res <- emptyDropsCellRanger(
    gex.mat,
    n.expected.cells = sum(sapply(md, function(x) x$cells_loaded)),
    umi.min = params$emptydrops_umi_min,
    umi.min.frac.median = params$emptydrops_umi_min_frac_median,
    cand.max.n = params$emptydrops_cand_max_n,
    ind.min = params$emptydrops_ind_min,
    ind.max = params$emptydrops_ind_max,
    niters = params$emptydrops_niters
  )
  bc.empty <- bc$gex[which(cc.res$FDR >= 0.001 | is.na(cc.res$FDR))]
  cc.stats <- list(
    cells = sum(cc.res$FDR < 0.001, na.rm = TRUE),
    empty = length(bc.empty),
    limited = any(cc.res$Limited[which(cc.res$FDR >= 0.001)])
  )
  bc$gex <- bc$gex[which(cc.res$FDR < 0.001)]
}
bcrank.res <- barcodeRanks(gex.mat)
bc.cells <- Reduce(intersect, bc)
gex.mat <- gex.mat[, bc.cells]
```

```{r cell-calling-classification, eval=!params$pre_filtered}
cat("### Classification", "\n\n")
cat("* **Cell-containing droplets:**", cc.stats$cells %>% format(big.mark = ","), "\n")
cat("* **Empty droplets:**", cc.stats$empty %>% format(big.mark = ","), "\n")
cat("\n\n")
ggplotly(
  bcrank.plot(
    bcrank.res,
    cells = bc.cells,
    title = "Barcode rank plot",
    x.lab = "Rank",
    y.lab = "Total UMI count",
    colour.lab = "Fraction cells",
    log = TRUE
  ) +
    scale_colour_gradient2(low = "#d82526", mid = "#ffc156", high = "#69b764", midpoint = 0.5)
)
```

```{r cell-calling-multimodal-plots, eval=all(!params$pre_filtered, length(modalities) > 1)}
cat("### Multimodal plots", "\n\n")
for (comb in combn(modalities, 2, FUN = c, simplify = FALSE)) {
  plot <- scatter.plot(
    data = total.counts %>%
      mutate(type = if_else(barcode %in% bc.cells, "Cell", "Empty")) %>%
      arrange(type) %>%
      select(all_of(comb), type) %>%
      distinct(),
    x = !!sym(comb[1]),
    y = !!sym(comb[2]),
    colour = type,
    size = 0.1,
    title = paste(toupper(comb[1]), "vs", toupper(comb[2]), sep = " "),
    x.lab = paste(toupper(comb[1]), if_else(comb[1] == "atac", "fragment", "UMI"), "count", sep = " "),
    y.lab = paste(toupper(comb[2]), if_else(comb[2] == "atac", "fragment", "UMI"), "count", sep = " "),
    log = TRUE
  ) +
    scale_colour_manual(values = list(Cell = "#69b764", Empty = "lightgrey"))
  print(plot)
}
```

```{r cell-calling-diagnostics, eval=!params$pre_filtered}
cat("### Diagnostics", "\n\n")
cat("* **Algorithm:**", params$cell_calling_algorithm, "\n")
if (params$cell_calling_algorithm == "joint") {
  cat("* **Modalities:**", params$joint_modalities[match(modalities, params$joint_modalities)] %>% toupper() %>% paste(collapse = ", "), "\n")
  cat("* **OrdMag quantile:**", params$joint_ordmag_quantile %>% format(big.mark = ","), "\n")
  cat("* **OrdMag ratio:**", params$joint_ordmag_ratio %>% format(big.mark = ","), "\n")
} else {
  cat("* **Minimum UMI count (absolute):**", params$emptydrops_umi_min %>% format(big.mark = ","), "\n")
  cat("* **Minimum UMI count (fraction of median):**", params$emptydrops_umi_min_frac_median %>% format(big.mark = ","), "\n")
  cat("* **Maximum number of candidate cells:**", params$emptydrops_cand_max_n %>% format(big.mark = ","), "\n")
  cat("* **Ambient estimation range:**", params$emptydrops_ind_min %>% format(big.mark = ","), "-", params$emptydrops_ind_max %>% format(big.mark = ","), "\n")
  cat("* **Monte Carlo simulation iterations:**", params$emptydrops_niters %>% format(big.mark = ","), "\n")
  cat("* **Limited:**", ifelse(cc.stats$limited, "Yes", "No"), "\n")
}
```

```{r hto-demultiplexing, eval=!is.null(params$hto_matrix)}
cat("## HTO demultiplexing {.tabset}", "\n\n")
set.seed(100)
hto.mat <- hto.mat[, bc.cells]
dmm <- demuxmix(
  hto = as.matrix(hto.mat),
  rna = colSums(gex.mat),
  model = params$demuxmix_model,
  pAcpt = params$demuxmix_pAcpt^length(wl)
)
models <- lapply(dmm@models, function(x) class(x))
demux.res <- dmm %>%
  dmmClassify() %>%
  mutate(Sample = if_else(grepl(",", HTO), "multiplet", HTO))
demux.stats <- demux.res %>%
  as.data.frame() %>%
  group_by(Sample) %>%
  summarise(count = n())
demux.stats <- demux.stats$count %>%
  as.list() %>%
  setNames(demux.stats$Sample)
```

```{r hto-demultiplexing-classification, eval=!is.null(params$hto_matrix)}
cat("### Classification", "\n\n")
cat(sprintf("* **%s:** %s", names(demux.stats), format(demux.stats, big.mark = ",")) %>% paste(collapse = "\n"), "\n")
cat("\n\n")
ggplotly(
  bar.plot(
    data = demux.res %>% mutate(Type = factor(Type, levels = c("singlet", "multiplet", "negative", "uncertain"))),
    pos = Type,
    fill = Sample,
    position = position_stack(reverse = TRUE),
    title = "Sample assignment",
    pos.lab = "Type",
    fill.lab = "Sample"
  ),
  tooltip = c("fill", "x")
)
```

```{r hto-demultiplexing-raw-counts, eval=!is.null(params$hto_matrix), fig.width=7, fig.height=length(wl)}
cat("### Raw HTO counts", "\n\n")
logcounts.plot(
  hto.mat,
  scale = 0.95,
  colour = NA,
  fill = "#595959",
  title = "Raw HTO counts"
)
```

```{r hto-demultiplexing-demux-counts, eval=!is.null(params$hto_matrix), fig.width=7, fig.height=3*ceiling(length(wl)/2)}
cat("### Demultiplexed HTO counts", "\n\n")
logcounts.plot(
  hto.mat,
  samples = demux.res$Sample,
  scale = 0.95,
  colour = NA,
  show.legend = FALSE,
  title = "Demultiplexed HTO counts"
)
```

```{r hto-demultiplexing-mixture-components, eval=!is.null(params$hto_matrix), fig.width=7, fig.height=2.5}
cat("### Mixture components", "\n\n")
for (x in wl) {
  plot <- plotDmmHistogram(dmm, hto = x) +
    labs(title = x, x = "UMI count") +
    theme_minimal()
  print(plot)
}
```

```{r hto-demultiplexing-diagnostics, eval=!is.null(params$hto_matrix)}
cat("### Diagnostics", "\n\n")
cat("* **Mode:**", params$demuxmix_model, "\n")
cat("* **Chosen model:**", "\n")
cat(sprintf("  + **%s:** %s", wl, models[wl]) %>% paste(collapse = "\n"), "\n")
cat("* **Acceptance probability:**", params$demuxmix_pAcpt^length(wl) %>% round(2), paste0("(", params$demuxmix_pAcpt %>% round(2), " per HTO)"), "\n")
cat("* **Fraction overlap between mixture components:**", "\n")
cat(sprintf("  + **%s:** %s", wl, round(dmmOverlap(dmm)[wl], 3)) %>% paste(collapse = "\n"), "\n")
```

```{r multiplet-calling}
cat("## Multiplet calling {.tabset}", "\n\n")
set.seed(100)
if (exists("demux.res")) {
  counts <- gex.mat[, demux.res$Type %in% c("singlet", "multiplet")]
  known <- demux.res[colnames(counts), "Type"] == "multiplet"
} else {
  counts <- gex.mat
  known <- rep(FALSE, times = ncol(counts))
}
sce <- SingleCellExperiment(assays = list(counts = counts)) %>% logNormCounts()
dec <- modelGeneVar(sce)
hvg <- getTopHVGs(dec, n = 1000)
sce <- sce %>% runPCA(subset_row = hvg, ncomponents = 20)
if (any(known)) {
  # If known inter-sample multiplets available, use nearest neighbours-based intra-sample multiplet prediction
  multiplet.res <- recoverDoublets(
    x = sce,
    use.dimred = "PCA",
    doublets = known,
    samples = table(demux.res$Sample[demux.res$Type %in% c("singlet", "multiplet")]) %>% head(-1)
  ) %>%
    as.data.frame() %>%
    select(known, predicted)
  rownames(multiplet.res) <- colnames(sce)
} else {
  # Otherwise, use simulation-based multiplet prediction
  multiplet.res <- scDblFinder(
    sce = sce,
    nfeatures = hvg,
    clusters = params$scdblfinder_clusters,
    dbr.sd = params$scdblfinder_dbr_sd,
    returnType = "table"
  ) %>%
    as.data.frame() %>%
    filter(type == "real")
  multiplet.res <- multiplet.res[colnames(sce), ]
  multiplet.res <- transmute(
    multiplet.res,
    known = known,
    predicted = if_else(class == "doublet", TRUE, FALSE)
  )
}
multiplet.stats <- list(
  known = sum(multiplet.res$known, na.rm = TRUE),
  predicted = sum(multiplet.res$predicted, na.rm = TRUE),
  singlets = sum(!multiplet.res$known & !multiplet.res$predicted, na.rm = TRUE)
)
```

```{r multiplet-calling-classification}
cat("### Classification", "\n\n")
cat("* **Singlets:**", multiplet.stats$singlets %>% format(big.mark = ","), "\n")
cat("* **Known (inter-sample) multiplets:**", multiplet.stats$known %>% format(big.mark = ","), "\n")
cat("* **Predicted (intra-sample) multiplets:**", multiplet.stats$predicted %>% format(big.mark = ","), "\n")
cat("\n\n")
ggplotly(
  multiplet.plot(
    sce,
    multiplet.res,
    size = 0.1,
    projection = params$multiplet_projection,
    use.dimred = "PCA",
    random.seed = 100
  ),
  tooltip = "colour"
)
```

# Output

```{r output}
metadata <- multiplet.res %>%
  filter(!known & !predicted) %>%
  transmute(
    Donor = stringr::str_split_i(params$sample_id, pattern = "-", i = 1),
    Pool = stringr::str_split_i(params$sample_id, pattern = "-", i = 2)
  ) %>%
  tibble::rownames_to_column("Barcode")
if (exists("demux.res")) metadata$Sample <- sapply(demux.res[metadata$Barcode, ]$Sample, function(x) md[[x]]$sample)
if (!dir.exists(params$outdir)) dir.create(params$outdir, recursive = TRUE)
vroom::vroom_write(
  metadata,
  file = file.path(params$outdir, "metadata.tsv"),
  delim = "\t"
)
if (exists("bc.empty")) {
  vroom::vroom_write_lines(
    bc.empty,
    file = file.path(params$outdir, "empty_droplet_barcodes.txt")
  )
}
cc.excluded <- if_else(exists("cc.stats"), cc.stats$empty, NA)
demux.excluded <- if_else(exists("demux.stats"), sum(demux.stats$negative, demux.stats$uncertain, na.rm = TRUE), NA)
cat("* **Total barcodes passed:**", multiplet.stats$singlets %>% format(big.mark = ","), "\n")
cat("* **Total barcodes excluded:**", sum(cc.excluded, demux.excluded, multiplet.stats$known, multiplet.stats$predicted, na.rm = TRUE) %>% format(big.mark = ","), "\n")
if (!is.na(cc.excluded)) cat("  + **Empty droplets:**", cc.excluded %>% format(big.mark = ","), "\n")
if (!is.na(demux.excluded)) cat("  + **Negative/uncertain HTO:**", demux.excluded %>% format(big.mark = ","), "\n")
cat("  + **Multiplets:**", sum(multiplet.stats$known, multiplet.stats$predicted, na.rm = TRUE) %>% format(big.mark = ","), "\n")
cat("\n\n")
metadata
```

# Session Info

```{r session-info, results="markup"}
sessionInfo()
```

# Source
This report was automatically generated from an R Markdown script (<a download="droplet_qc.Rmd" href="`r base64enc::dataURI(file = params$rmd, mime = 'text/rmd', encoding = 'base64')`">download source</a>).
